Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
table
paper s5 conflics solved

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. 
\MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
\EKHINE{es decreciente igual}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

@@

\myparagraph{Comparison of UMM and CEGO}

When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

\myparagraph{Comparison of the best solutions found} We now turn our analysis
to the best (minimum) fitness found at the end of the run by each algorithm. We
show in Table~\ref{tab:results} the mean fitness of the best solution found
over the 10 runs of each algorithm.

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found. % at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of UMM minus the one of CEGO. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
  \resizebox{\textwidth}{!}{%
\begin{tabular}{r*{5}{r}r@{\hskip 0.5ex}l*{2}{r}}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-p50-02     & 19090.5                      & (285.4)                                                                         & 21746.6   & (645.4)   & {[}-3140.6,& -2171.6{]}    & 22.9 & 0.5 \\
         & N-p50-01     & 18197.7                      & (225.6)                                                                         & 21161.8   & (657.4)   & {[}-3447.4,& -2480.8{]}    & 23.2 & 0.5 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}}
\end{table*}

>>>>

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. 
\MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
\EKHINE{es decreciente igual}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values.

%The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

@@

\myparagraph{Comparison of UMM and CEGO}
%
When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show in
Table~\ref{tab:results} the mean fitness (and standard deviation) of the best
solution found over the 10 runs of each algorithm, together with a 95\%
confidence interval (CI) around their mean fitness difference. Except for PFSP
instances \texttt{rec05} and \texttt{rec31}, the differences are statistically
significant in favour of \CEGOinv (the intervals do not contain the value
zero). Nevertheless, the CIs indicate that the differences, although
consistent, are relatively small.

\myparagraph{Runtime comparison}
%
Table~\ref{tab:results} also shows the mean
runtime of \CEGOinv and UMM$_{400}$ in hours. The difference in computational
effort between \CEGOinv and UMM is at least two orders of magnitude.

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-p50-02     & 19090.5                      & (285.4)                                                                         & 21746.6   & (645.4)   & {[}-3140.6,& -2171.6{]}    & 22.9 & 0.5 \\
         & N-p50-01     & 18197.7                      & (225.6)                                                                         & 21161.8   & (657.4)   & {[}-3447.4,& -2480.8{]}    & 23.2 & 0.5 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}
